Follow-up Samples

### 2016-12-16 submission/manuscript_outline_20161024.tex
update introduction Pat's suggestions

Colorectal cancer (CRC) continues to be a leading cause of cancer
related deaths and is the second most common cancer death among men aged
40-79 years of age (1, 2). Over the last few years death due to the
disease has seen a significant decrease thanks mainly to improvements in
screening (1). However, despite this giant improvement there are still
approximately 50,000 deaths from the disease a year (2). It is estimated
that around 5-10\% of all CRCs can be explained by autosomal dominant
inheritance (3). The vast majority of CRCs are not inherited and the
exact etiology to disease has not been well worked out (2).

Recent reports have found that the microbiome between those with and
without CRC are different and that bacteia associated with CRC tissue is
different than that found on the normal mucosa (\textbf{Insert
citations}). Further, there is evidence to suggest that more
inflammatory mouth-associated bacteria are more prevalent in the guts of
those with CRC (\textbf{insert citation}). This has led to the
hypothesis that these bacteria replace traditional residents that
produce metabolites such as butyrate (\textbf{insert citation}) and lead
to a more inflammatory state (\textbf{insert citation}). This
inflammatory state allows for more of these types of bacteria to
populate the gut and leads to neoplasia and eventually CRC
(\textbf{insert citation}).

Building upon these findings a number of groups, including our own, have
investigated using the bacterial microbiome as a potential biomarker for
adenoma and CRC detection (\textbf{insert citations}). These
collectively show that a reasonable area under the curve (AUC) can be
obtained when using the stool bacterial microbiome to classify disease
state (\textbf{insert citation}). Further, a number of groups have
extended this observation to show that using either specific bacterial
species or the bacterial microbiome as a whole paired with the Fecal
Immunochemical Test (FIT) can increase the AUC of the model
(\textbf{insert citations}).

In this study we further refine the random forest model by including
more background information on the individuals within it. We used a
rigourous and accepted approach for machine learning training and
testing to validate this model (\textbf{insert citation}). We then
applied it to individuals in which we have samples before and after
treatment to assess how well the model adjusts their probability of
having an adenoma or carcinoma. Finally, we investigated which specific
Operational Taxonomic Units (OTUs) are most affected by treatment.
Within this study lesion refers to both adenoma and carcinoma.

@@

>>>>

Colorectal cancer (CRC) continues to be a leading cause of cancer
related deaths and is the second most common cancer death among men aged
40-79 years of age (1, 2). Over the last few years death due to the
disease has seen a significant decrease thanks mainly to improvements in
screening (1). However, despite this giant improvement there are still
approximately 50,000 deaths from the disease a year (2). It is estimated
that around 5-10\% of all CRCs can be explained by autosomal dominant
inheritance (3). The vast majority of CRCs are not inherited and the
exact etiology to disease has not been well worked out (2). Although
many risk factors have been identified ({\textbf{???}}) and non-invasive
screening techniques have started to be put into consistent use (4, 5)
there has been a consistent increase in the incidence of CRC in the
younger population.

This increased incidence of CRC in the younger population is concerncing
since having either an adenoma or carcinoma increases ones risk for
future adenomas or carcinomas (6--8). This increased risk can also carry
with it an increased risk of mortality due to this recurrence (9, 10).
Therefore there has been a great amount of interest in early risk
stratification tools (11, 12) that can help identify those they may be
at risk of reccurence. Concurrently with this there has been a lot of
interest in new areas such as the gut bacterial microbiome for insight
into potential disease pathology.

There has been promising work on the bacterial microbiome and it's
ability to be able to complement existing screening methods such as
Fecal Immunoglobulin Test (FIT) or act alone as a screening tool (13,
14). There has also been research into how this microbiome could be
altered directly on tumor tissue itself (15). There have also been a few
studies that have shown how this microbiome (16) or specific members
within it (17) could be directly involved ith the pathogenesis of CRC.
These studies have helped to provide a tantilzing link between the
bacterial microbiome and CRC. However, at this present time there
remains limited information on the bacterial microbiome before and after
successful surgery for removal of the adenoma or carcinoma.

In this study we investigated what happened to the bacterial microbiome
before and after surgery for both adenoma and carcinoma individuals. Our
anlaysis includes both alpha and beta diversity analysis along with
investigating indvidual operational taxonomic units (OTUs). We next
utilized a Random Forest model that was trained on a completely separate
data set and observed how this model as well as specific OTUs within
this model performed on the initial and follow up samples. We then
created a new model to classify intial and follow up samples and
investigated how the model performs and what specific OUTs within the
model change. Finally, we observed how specific bacteria that have been
previously implicated in CRC changed before and after treatment in both
the adenoma and carcinoma individuals.

@@

\hypertarget{ref-liao_application_2013}{}
4. \textbf{Liao C-S}, \textbf{Lin Y-M}, \textbf{Chang H-C}, \textbf{Chen
Y-H}, \textbf{Chong L-W}, \textbf{Chen C-H}, \textbf{Lin Y-S},
\textbf{Yang K-C}, \textbf{Shih C-H}. 2013. Application of quantitative
estimates of fecal hemoglobin concentration for risk prediction of
colorectal neoplasia. World Journal of Gastroenterology
\textbf{19}:8366--8372.
doi:\href{https://doi.org/10.3748/wjg.v19.i45.8366}{10.3748/wjg.v19.i45.8366}.

\hypertarget{ref-johnson_multi-target_2016}{}
5. \textbf{Johnson DH}, \textbf{Kisiel JB}, \textbf{Burger KN},
\textbf{Mahoney DW}, \textbf{Devens ME}, \textbf{Ahlquist DA},
\textbf{Sweetser S}. 2016. Multi-target stool DNA test: Clinical
performance and impact on yield and quality of colonoscopy for
colorectal cancer screening. Gastrointestinal Endoscopy.
doi:\href{https://doi.org/10.1016/j.gie.2016.11.012}{10.1016/j.gie.2016.11.012}.

\hypertarget{ref-laiyemo_short-_2013}{}
6. \textbf{Laiyemo AO}, \textbf{Doubeni C}, \textbf{Brim H},
\textbf{Ashktorab H}, \textbf{Schoen RE}, \textbf{Gupta S},
\textbf{Charabaty A}, \textbf{Lanza E}, \textbf{Smoot DT}, \textbf{Platz
E}, \textbf{Cross AJ}. 2013. Short- and long-term risk of colorectal
adenoma recurrence among whites and blacks. Gastrointestinal Endoscopy
\textbf{77}:447--454.
doi:\href{https://doi.org/10.1016/j.gie.2012.11.027}{10.1016/j.gie.2012.11.027}.

\hypertarget{ref-matsuda_five-year_2009}{}
7. \textbf{Matsuda T}, \textbf{Fujii T}, \textbf{Sano Y}, \textbf{Kudo
S-e}, \textbf{Oda Y}, \textbf{Igarashi M}, \textbf{Iishi H},
\textbf{Murakami Y}, \textbf{Ishikawa H}, \textbf{Shimoda T},
\textbf{Kaneko K}, \textbf{Yoshida S}. 2009. Five-year incidence of
advanced neoplasia after initial colonoscopy in Japan: A multicenter
retrospective cohort study. Japanese Journal of Clinical Oncology
\textbf{39}:435--442.
doi:\href{https://doi.org/10.1093/jjco/hyp047}{10.1093/jjco/hyp047}.

\hypertarget{ref-ren_long-term_2016}{}
8. \textbf{Ren J}, \textbf{Kirkness CS}, \textbf{Kim M}, \textbf{Asche
CV}, \textbf{Puli S}. 2016. Long-term risk of colorectal cancer by
gender after positive colonoscopy: Population-based cohort study.
Current Medical Research and Opinion \textbf{32}:1367--1374.
doi:\href{https://doi.org/10.1080/03007995.2016.1174840}{10.1080/03007995.2016.1174840}.

\hypertarget{ref-loberg_long-term_2014}{}
9. \textbf{Løberg M}, \textbf{Kalager M}, \textbf{Holme Ø}, \textbf{Hoff
G}, \textbf{Adami H-O}, \textbf{Bretthauer M}. 2014. Long-term
colorectal-cancer mortality after adenoma removal. The New England
Journal of Medicine \textbf{371}:799--807.
doi:\href{https://doi.org/10.1056/NEJMoa1315870}{10.1056/NEJMoa1315870}.

\hypertarget{ref-freeman_natural_2013}{}
10. \textbf{Freeman HJ}. 2013. Natural history and long-term outcomes of
patients treated for early stage colorectal cancer. Canadian Journal of
Gastroenterology = Journal Canadien De Gastroenterologie
\textbf{27}:409--413.

\hypertarget{ref-lee_identification_2016}{}
11. \textbf{Lee JH}, \textbf{Lee JL}, \textbf{Park IJ}, \textbf{Lim
S-B}, \textbf{Yu CS}, \textbf{Kim JC}. 2016. Identification of
Recurrence-Predictive Indicators in Stage I Colorectal Cancer. World
Journal of Surgery.
doi:\href{https://doi.org/10.1007/s00268-016-3833-2}{10.1007/s00268-016-3833-2}.

\hypertarget{ref-richards_evidence-based_2016}{}
12. \textbf{Richards CH}, \textbf{Ventham NT}, \textbf{Mansouri D},
\textbf{Wilson M}, \textbf{Ramsay G}, \textbf{Mackay CD},
\textbf{Parnaby CN}, \textbf{Smith D}, \textbf{On J}, \textbf{Speake D},
\textbf{McFarlane G}, \textbf{Neo YN}, \textbf{Aitken E},
\textbf{Forrest C}, \textbf{Knight K}, \textbf{McKay A}, \textbf{Nair
H}, \textbf{Mulholland C}, \textbf{Robertson JH}, \textbf{Carey FA},
\textbf{Steele R}, \textbf{Scottish Surgical Research Group}. 2016. An
evidence-based treatment algorithm for colorectal polyp cancers: Results
from the Scottish Screen-detected Polyp Cancer Study (SSPoCS). Gut.
doi:\href{https://doi.org/10.1136/gutjnl-2016-312201}{10.1136/gutjnl-2016-312201}.

\hypertarget{ref-baxter_microbiota-based_2016}{}
13. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
\textbf{Schloss PD}. 2016. Microbiota-based model improves the
sensitivity of fecal immunochemical test for detecting colonic lesions.
Genome Medicine \textbf{8}:37.
doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.

\hypertarget{ref-zeller_potential_2014}{}
14. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
\textbf{Sunagawa S}, \textbf{Kultima JR}, \textbf{Costea PI},
\textbf{Amiot A}, \textbf{Böhm J}, \textbf{Brunetti F},
\textbf{Habermann N}, \textbf{Hercog R}, \textbf{Koch M},
\textbf{Luciani A}, \textbf{Mende DR}, \textbf{Schneider MA},
\textbf{Schrotz-King P}, \textbf{Tournigand C}, \textbf{Tran Van Nhieu
J}, \textbf{Yamada T}, \textbf{Zimmermann J}, \textbf{Benes V},
\textbf{Kloor M}, \textbf{Ulrich CM}, \textbf{Knebel Doeberitz M von},
\textbf{Sobhani I}, \textbf{Bork P}. 2014. Potential of fecal microbiota
for early-stage detection of colorectal cancer. Molecular Systems
Biology \textbf{10}:766.

\hypertarget{ref-dejea_microbiota_2014}{}
15. \textbf{Dejea CM}, \textbf{Wick EC}, \textbf{Hechenbleikner EM},
\textbf{White JR}, \textbf{Mark Welch JL}, \textbf{Rossetti BJ},
\textbf{Peterson SN}, \textbf{Snesrud EC}, \textbf{Borisy GG},
\textbf{Lazarev M}, \textbf{Stein E}, \textbf{Vadivelu J},
\textbf{Roslani AC}, \textbf{Malik AA}, \textbf{Wanyiri JW}, \textbf{Goh
KL}, \textbf{Thevambiga I}, \textbf{Fu K}, \textbf{Wan F}, \textbf{Llosa
N}, \textbf{Housseau F}, \textbf{Romans K}, \textbf{Wu X},
\textbf{McAllister FM}, \textbf{Wu S}, \textbf{Vogelstein B},
\textbf{Kinzler KW}, \textbf{Pardoll DM}, \textbf{Sears CL}. 2014.
Microbiota organization is a distinct feature of proximal colorectal
cancers. Proceedings of the National Academy of Sciences of the United
States of America \textbf{111}:18321--18326.
doi:\href{https://doi.org/10.1073/pnas.1406199111}{10.1073/pnas.1406199111}.

\hypertarget{ref-zackular_manipulation_2016}{}
16. \textbf{Zackular JP}, \textbf{Baxter NT}, \textbf{Chen GY},
\textbf{Schloss PD}. 2016. Manipulation of the Gut Microbiota Reveals
Role in Colon Tumorigenesis. mSphere \textbf{1}.
doi:\href{https://doi.org/10.1128/mSphere.00001-15}{10.1128/mSphere.00001-15}.

\hypertarget{ref-arthur_microbial_2014}{}
17. \textbf{Arthur JC}, \textbf{Gharaibeh RZ}, \textbf{Mühlbauer M},
\textbf{Perez-Chanona E}, \textbf{Uronis JM}, \textbf{McCafferty J},
\textbf{Fodor AA}, \textbf{Jobin C}. 2014. Microbial genomic analysis
reveals the essential role of inflammation in bacteria-induced
colorectal cancer. Nature Communications \textbf{5}:4724.
doi:\href{https://doi.org/10.1038/ncomms5724}{10.1038/ncomms5724}.
